---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ruODK
[Code on GitHub](https://github.com/dbca-wa/ruODK)

Especially in these trying times, it is important to ask: "R U ODK?"

This package aims to provide support, and with that we mean technical support, 
to the data wranglers trying to get data out of ODK's new data warehouse, ODK Central.

ODK Central OData API specs are [here](https://odkcentral.docs.apiary.io/#reference/odata-endpoints).

## Installation

You can install ruODK from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("dbca-wa/ruODK")
```


## Use

### Setup ODK Central

The ODK Central [user manual](https://docs.opendatakit.org/central-using/) 
provides up-to-date descriptions of the steps below.

* [Create a web user account](https://docs.opendatakit.org/central-users/#creating-a-web-user) 
  on an ODK Central instance. Your username will be an email address.
* Create a [project](https://docs.opendatakit.org/central-projects/) and give the
  web user the relevant permissions.
* Create a Xform, e.g. using ODK Build, or use the provided example forms.
* Publish the [form to ODK Central](https://docs.opendatakit.org/central-forms/).
* Collect data for this form on ODK Collect.
* Get the ODK Central OData service URL.

A note on the [included example forms](https://github.com/dbca-wa/ruODK/tree/master/inst/extdata): 
The `.odkbuild` versions can be loaded into [ODK Build](https://build.opendatakit.org/), 
while the `.xml` versions can be imported into ODK Central.

### Configure ruODK

* Set your ODK Central username (email) and password as R environment
  variables, e.g. in your `~/.Rprofile`.
  Example:
  
```{r, eval=F}
Sys.setenv(ODKC_UN="...@...")
Sys.setenv(ODKC_PW=".......")
```

### Use ruODK

ruODK aims to provide the same access to your submissions via ODK Central's OData
service endpoints as MS Excel, MS PowerBI and Tableau.

At the end of this step, we want to achieve the same outcome as the [manual
download to CSV](https://docs.opendatakit.org/central-submissions/#downloading-submissions-as-csvs).

Caveat: this is a work in progress. Some data doesn't come through the OData feed
(such as location and altitude accuracy).

An [example](https://rpubs.com/florian_mayer/flora_quadrats):

```{r, eval=F}
# ODK Central credentials
if (file.exists("~/.Rprofile")) source("~/.Rprofile")

# ODK Central
data_url <- "https://sandbox.central.opendatakit.org/v1/projects/14/forms/build_Flora-Quadrat-0-2_1558575936.svc"

# Download from ODK Central
metadata <- get_metadata(data_url)
data <- data_url %>% get_submissions() %>% parse_submissions()
```

See `vignette("example")` for a walk-through with some data visualisation.


## Contribute
Contributions through issues and PRs are welcome!

## Acknowledgements
The Department of Biodiversity, Conservation and Attractions (DBCA) respectfully acknowledges Aboriginal people as the traditional owners of the lands and waters it manages.

One of the Department's core missions is to conserve and protect the value of the land to the culture and heritage of Aboriginal people.

This software was created both as a contribution to the ODK ecosystem and for the conservation of the biodiversity of Western Australia, and in doing so, caring for country.