---
title: "Getting data from the API"
author: "Florian Mayer, DBCA"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{restapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ruODK)
```

# Scope
This vignette provides a walk-through of the "getting data out" functions of
the RESTful API endpoints which list and view details.
This is useful to build ETL pipelines from ODK Central into other data warehouses,
or to build automated and reproducible reporting. 

The result will be similar to

* browsing ODK Central and downloading submissions as zip archive,
* viewing ODK Central data in MS PowerBI, MS Excel, or Tableau through the
  OData service, or
* viewing ODK Central data through ruODK's OData functions.

Not implemented (yet) are the "managing ODK Central" functions which create, 
update, and delete projects, forms, users, roles, and permissions. 
We haven't yet found a strong use case to automate those functions - 
ODK Central (driven by humans) does those jobs beautifully.

# Setup ruODK
See [`vignette("Setup", package = "ruODK")`](https://dbca-wa.github.io/ruODK/articles/setup.html) 
for detailed options to configure `ruODK`.

# Projects

List projects.

```{r project_list}
p <- project_list()
p %>% knitr::kable(.)
```

Inspect a project. Turtles are, not surprisingly, number 1. Of course they are.

```{r project_details}
pd <- project_detail(1)
pd %>% knitr::kable(.)
```

# Forms

```{r form_list}
forms <- form_list(1)
forms  %>% dplyr::select(-"xml") %>% knitr::kable(.)
```

```{r form_detail}
form <- form_detail(1, forms$xmlFormId[[1]])
form %>% dplyr::select(-"xml") %>% knitr::kable(.)
```

# Submissions
Coming soon

# Attachments
Comin soon