---
title: "Accessing the RESTful API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{restapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ruODK)
```

# Scope
This vignette provides a walk-through of the "getting data out" 
functions of the RESTful API endpoints which list and view details.

`ruODK` users would mix and match parts of the demonstrated workflows to build
their own data pipelines, e.g.:

* to build a quick analysis from all data, freshly downloaded from a smaller 
  project, or
* to build an interactive ETL pipeline to selectively download only new submissions
  for further processing and upload into downstream data warehouses.

## Three ways to happiness

ODK Central offers no less than three different ways to access data:

* viewing ODK Central data in MS PowerBI, MS Excel, Tableau, or `ruODK` 
  through the OData service endpoints, or
* downloading all submissions including attachments as one (possibly gigantic) 
  zip archive either through the "Export Submissions" button in the ODK Central
  form submissions page or through `ruODK`,
* viewing ODK Central data through `ruODK`'s RESTful API functions.

While the `vignette("odata", package="ruODK")` 
(online [here](https://dbca-wa.github.io/ruODK/articles/odata.html)) illustrates 
the first option, this vignette demonstrates the remaining two.

Not implemented (yet) are the "managing ODK Central" functions which create, 
update, and delete projects, forms, users, roles, and permissions. 
We haven't yet found a strong use case to automate those functions - 
ODK Central (driven by humans) does those jobs beautifully on an expected scale.

# Setup ruODK
See [`vignette("Setup", package = "ruODK")`](https://dbca-wa.github.io/ruODK/articles/setup.html) 
for detailed options to configure `ruODK`.

Here, we'll grab the OData service URL from the form used in this vignette,
plus username and password of a web user of ODK Central with access to that form.

`ruODK::ru_setup()` will populate the default url, project ID, and form ID which
are used by `ruODK`'s other functions (unless specified otherwise).

```{r ru_setup}
ruODK::ru_setup(
  svc = "https://sandbox.central.opendatakit.org/v1/projects/14/forms/build_Flora-Quadrat-0-2_1558575936.svc",
  un = Sys.getenv("ODKC_TEST_UN"),
  pw = Sys.getenv("ODKC_TEST_PW")
)
```

# Projects

List projects. We see the project ID, a name, the number of forms and app users,
dates of last form submissions plus project management timestamps (created, 
updated).

The important bit here is the project ID.

```{r project_list}
p <- ruODK::project_list()
p %>% knitr::kable(.)
```

Inspect a project using its ID. We receive a tibble with exactly one row,
all the columns of `ruODK::project_list` plus a column `verbs`, which contains all
available API actions for a project.

```{r project_details}
pd <- ruODK::project_detail()

# Project details (without verbs)
pd %>% dplyr::select(-"verbs") %>%  knitr::kable(.)

# Available verbs
pd$verbs[[1]] %>% unlist(.)
```

Nothing apart from the verbs is new compared to the data returned by 
`ruODK::project_list`.

To learn more about the functionality behind the verbs, refer to the interactive 
[ODK Central API documentation](https://odkcentral.docs.apiary.io/#reference/project-management).

To retrieve data from ODK Central, the functions shown below will suffice.

# Forms
## List forms for a project

To download form submissions, we need to know project ID and form ID.

There are several ways of retrieving the form ID:

* Browsing forms in the ODK Central's project overviews,
* Stealing the form ID from the OData service endpoint URL as shown on 
  ODK Central's form submission page,
* Listing form metadata for a given project ID with `ruODK::form_list()`.

```{r form_list}
forms <- ruODK::form_list() 
forms %>% knitr::kable(.)
```

Further to the metadata shown here, a column `xml` contains the entire XForms
definition (originally XML) as nested list.

If the original XML is needed rather than the R equivalent (nested list), 
we can use `ruODK::form_xml` with parameter `parse=FALSE`:

```{r form_xml}
form_xml <- ruODK::form_xml(parse=FALSE)
form_xml
```

## Inspect form schema

The nested list of lists returned by `form_schema` is the cleanest representation
of the XForm definition available in R.

See the [ODK Central API docs](https://odkcentral.docs.apiary.io/#reference/forms-and-submissions/'-individual-form/retrieving-form-schema-json) and the examples of  `??ruODK::form_schema()` for more detail.

```{r form_schema, fig.width=7}
fs <- form_schema()
listviewer::jsonedit(form_schema, mode="code")
```

The form schema tells us about form field data types (e.g. dates, geopoints, 
binary attachments) and repeating groups. We need this knowledge to parse the 
submission data into tidy tibbles later.

Feel free to search this list for "binary" to find the field names of file 
attachments. Similarly, search for "dateTime" to find field names of datetimes.

## Show details of one form

The details of a form are exactly the same as the output of `ruODK::form_list()`.

```{r form_detail}
form <- ruODK::form_detail()
form %>% knitr::kable(.)
```

# Submissions
We are getting closer to the actual data! This section shows two of the options
for data access: dump all submissions, or extract a subset.

## Get all submissions for one form
Smaller datasets lend themselves to be exported in one go.
ODK Central offers one giant zip file containing all submissions, any
repeating groups, and any attachments both on the form submission page, and as
API endpoint which is provided as `ruODK::submission_export()`.

The default behaviour of `ruODK::submission_export()` is to write the zip file
to the project root (`here::here()`), and to overwrite existing previous downloads.
See `?ruODK::submission_export()` for alternative download and retention options.

In the following chuck, we illustrate common tasks:

* Download the zip file.
* Unpack the zip file.
* Join repeating form group data `data_taxon` to main data `data_quadrat` to
  annotate `data_taxon` with data from `data_quadrat`.
* Sanitise the column names.
* Prepend all attachment filenames (e.g. `data_quadrat$location_quadrat_photo`,
  `data_taxon$photo_in_situ`) with `media/`.


```{r submission_export}
t <- fs::dir_create("attachments")

# Predict filenames (with knowledge of form)
fid <- ruODK::get_test_fid() # "build_Flora-Quadrat-0-2_1558575936.svc"
fid_csv <- fs::path(t, glue::glue("{fid}.csv"))
fid_csv_tae <- fs::path(t, glue::glue("{fid}-taxon_encounter.csv"))

# Download the zip file
se <- ruODK::submission_export(
  local_dir = t,
  overwrite = FALSE,
  verbose = TRUE
)

# Unpack the zip file
f <- unzip(se, exdir = t)
fs::dir_ls(t)

# Prepend attachments with media/ to turn into relative file paths
data_quadrat <- fid_csv %>% 
  readr::read_csv(na = c("", "NA", "na")) %>% # form uses "na" for NA
  janitor::clean_names(.) %>% 
  attachment_link(.) %>% 
  parse_datetime(tz = "Australia/Perth")      # an example timezone

DT::datatable(data_quadrat)
# Further: create map with popups, see vignette "odata"

data_taxon <- fid_csv_tae %>%
  readr::read_csv(na = c("", "NA", "na")) %>%
  janitor::clean_names(.) %>%
  attachment_link(.) %>%
  parse_datetime(tz = "Australia/Perth") %>%
  dplyr::left_join(data_quadrat, by = c("parent_key" = "meta_instance_id"))

DT::datatable(data_taxon)
# Further: create map with popups, see vignette "odata"
```

## List submissions for one form
Not always is it appropriate to download all submissions and all attachments
at once. 

If forms feed into downstream data warehouses, the typical ETL workflow is to 

* List all submissions from ODK Central
* Select the subset of new submissions to download, e.g.
  * Submissions younger than the oldest submission date in the data warehouse.
  * Submissions whose `instance_id` is not already present in the data warehouse.
* Download only the selected submissions.
* Download attachments of only the selected submissions.

```{r submission_list}
sl <- ruODK::submission_list()
sl %>% knitr::kable(.)
```

The list of submissions critically contains each submission's unique ID in
`instance_id`. If the submissions shall be downloaded and uploaded into another
data warehouse, the `instance_id` can be used to determine whether a record
already exists in the downstream warehouse or not.
This workflow is preferable where the majority of submissions is already 
imported into another downstream data warehouse, and we only want to add new 
submissions, as in submissions which are not already imported into the data 
warehouse.

Furthermore, the `instance_id`s can now be used to retrieve the actual 
submissions.

A quick detour on submission metadata:
```{r submission_metadata}
# One submission
sub_meta <- ruODK::submission_detail(sl$instance_id[[1]])

# Multiple submissions
submissions_meta <- tibble::tibble(
  pid = ruODK::get_test_pid(),
  fid = ruODK::get_test_fid(),
  iid = sl$instance_id,  # this is a vector of multiple instance_ids
  url = ruODK::get_test_url(),
  un = ruODK::get_test_un(),
  pw = ruODK::get_test_pw()
) %>% 
  purrr::pmap(ruODK::submission_detail)
submissions_meta %>% knitr::kable(.)
```

Nothing new there compared to `submission_list`, so let's move on to the actual
submission data.

## Get submission data

In order to import each submission, we need to retrieve the data by 
`instance_id`.

```{r submission_data}
# One submission
sub <- ruODK::submission_get(sl$instance_id[[1]])

# Multiple submissions
submissions <- tibble::tibble(
  pid = ruODK::get_test_pid(),
  fid = ruODK::get_test_fid(),
  iid = sl$instance_id,  # this is a vector of multiple instance_ids
  url = ruODK::get_test_url(),
  un = ruODK::get_test_un(),
  pw = ruODK::get_test_pw()
) %>% 
  purrr::pmap(ruODK::submission_get)
```

## Parse submissions
The data in `sub` is one row of the bulk downloaded submissions in `data_quadrat`.
The data in `submissions` represents all (or let's pretend, the selected) 
submissions in `data_quadrat`.
The field `xml` contains the actual submission data including repeating groups.

The structure is different to the output of `ruODK::odata_submission_get`,
therefore `ruODK::odata_submission_parse` does not work for those, as
here we might have repeating groups included in a submission.

```{r view_submission_data}
listviewer::jsonedit(submissions, mode = "code")
```

## Rectangling the submissions
Here we'll un-nest the nested list of submission data into a rectangular `tibble`.

In the example shown above, the repeated `taxon_encounter` groups are named
`taxon_encounter`, `taxon_encounter.1` and so on. This is an artifact of the 
`jsonedit::listviewer()` widget.

With one additional step we can flatten the nested lists and bind them into one
dataframe, then sanitise the column names and parse datetimes like we did with
the CSV export.

It appears this operation loses all but the last repeat of repeating groups.
We should further also deparse locations (WKT) into latitude, longitude, 
altitude and accuracy. 

```{r parse_submissions}
sub_parsed <- submissions %>% 
  purrr::map_df(rlist::list.flatten) %>%
  janitor::clean_names(.) %>% 
  ruODK::parse_datetime(.)

names(sub_parsed)
sub_parsed %>% knitr::kable(.)
```

# Attachments

## List all attachments for one form

```{r attachment_list}
al <- ruODK::attachment_list(sl$instance_id[[1]])
al

attachments <- tibble::tibble(
  pid = ruODK::get_test_pid(),
  fid = ruODK::get_test_fid(),
  iid = sl$instance_id,  # this is a vector of multiple instance_ids
  url = ruODK::get_test_url(),
  un = ruODK::get_test_un(),
  pw = ruODK::get_test_pw()
) %>% 
  purrr::pmap(ruODK::attachment_list)
attachments
```

## Get attachments
While we're working on parsing the submissions into something resembling
`data_quadrat`, we'll demonstrate the use of `ruODK::attachment_get` using
`data_quadrat_raw`, which still contains the plain filenames without prefixed 
local folders.

`ruODK::attachment_get` downloads the attachment files and links them similar
to `ruODK::attachment_link` (which assumes that attachments are already 
downloaded).

```{r attachment_get}
sub_att <- sub_parsed %>% 
  dplyr::mutate(
    location_quadrat_photo = ruODK::attachment_get(
        meta_instance_id,
        location_quadrat_photo,
        local_dir = fs::path(t, "media")
    )
    # same for all other attachments
    # requires knowledge or introspection of form schema
  )

# We have file paths...
sub_att$location_quadrat_photo

# ...and the corresponding files
fs::dir_ls(t, recurse = TRUE, glob = "*.jpg")
```

# Outlook

From here, we can use the separately created data tibbles and locally downloaded
attachments to build visualisations as shown in the vignette "odata".

The best course of action seems to be the data export to ZIP.

# Future development

In future versions of `ruODK`, we'll introspect `form_schema` to determine
the names of repeating groups, the form field names of attachments (data type 
"binary" in form schema), the form field names of date/time fields (data type 
"dateTime"), and use this introspection to further automate the data extraction
process shown above.
