---
title: "Example"
author: "Florian Mayer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(DT)
library(leaflet)
library(listviewer)
library(magrittr)
library(tibble)
# library(tidyr)
library(ruODK)
```

# Configure ruODK

Set your ODK Central username (email) and password as R environment variables, 
e.g. in your `~/.Rprofile`.

```{r sys_config, eval=F}
Sys.setenv(ODKC_UN="...@...")
Sys.setenv(ODKC_PW=".......")
```


# Read data from ODK Central
If we know our ODK Central URL, the project ID (a number) and the respective
form ID (an alphanumeric string), we can infer the OData service URL.

Alternatively, the OData service URL is shown in the form's "Submissions" tab >
"Analyze via OData" on ODK Central.

```{r odk_config}
# ODK Central credentials
if (file.exists("~/.Rprofile")) source("~/.Rprofile")

# ODK Central
odk_central <- "https://sandbox.central.opendatakit.org/"
project_id <- 14
form_id <- "build_Flora-Quadrat-0-2_1558575936"

base_url <- glue::glue("{odk_central}v1/projects/{project_id}/forms/")
data_url <- glue::glue("{base_url}{form_id}.svc")

data("fq_meta")
data("fq_raw")
data("fq_tae")
```

In our example, the OData service URL is `r data_url`.
Let's first get the metadata to discover our data schema.

```{r load_ometadata, eval=F}
fq_meta <- get_metadata(data_url)
```

Each downloadable table is an EntityType under DataServices.Schema.
The main form is called "Submissions". 
Repeating form groups, as in our example "taxon_encounter", are named "Submissions.taxon_encounter". The exact name is found under
DataServices.Schema.EntityType[n].attrs.Name.

```{r view_metadata}
listviewer::jsonedit(fq_meta)
```

So let's download the form submissions and, separately, repeating form groups.

```{r load_odata, eval=F}
fq_raw <- get_submissions(data_url)
fq_tae <- get_submissions(data_url, sub="Submissions.taxon_encounter")
```

The output of the above code is provided as data objects `fq_raw` 
(main submissions of form Flora Quadrat 0.2) and 
`fq_tae` (repeated group "Taxon Encounter" within a Flora Quadrat).

# Rectangle the data
The nested list of lists must be transformed into a rectangular shape.
A helper `ruODK::parse_submissions` recursively unnests list columns using
(tidyr::unnest_wider). Unnamed columns, prominently lat/lon/alt coordinates,
are named automatically to become unique. The result

The unnesting could also be done manually by building up a pipeline, which
stepwise unnests each list column.
This requires knowledge of the data structure, which can either be looked up
from the metadata, or by inspecting the raw data, `fq_raw`.

The repeated subgroup `taxon_encounter` is joined with the main observation
to receive a (repeated) copy of the main observation's values 
(such as observation time and habitat description).

The vectorised function `get_attachment` downloads and links attachments like 
photos and other media to a local, relative path.

The trailing `invisible` is a neutral operation to allow trailing magrittr 
pipelines `%>%` while commenting individual lines of the pipelines in or out.

```{r transform_data, message=F}
fq_data <- fq_raw %>% parse_submissions(debug=T) %>% 
  dplyr::rename(
    uuid=`.__id`,) %>% 
  dplyr::mutate(
    quadrat_photo = get_attachment
    (base_url, form_id, uuid, quadrat_photo),
    morphological_type_photo = get_attachment(
      base_url, form_id, uuid, morphological_type_photo),
    mudmap_photo = get_attachment(
      base_url, form_id, uuid, mudmap_photo)
  ) %>% 
  invisible

fq_data_tae <- fq_tae %>% 
  parse_submissions(debug=T) %>% 
  dplyr::rename(
    tae_uuid=`.__id`,
    uuid=`.__Submissions.id`) %>%
  dplyr::mutate(
    photo_in_situ = get_attachment(
      base_url, form_id, uuid, photo_in_situ)) %>% 
  dplyr::left_join(fq_data, by="uuid") %>% 
  invisible
```

Note: A manually resized version of the original photos in this example live in `docs/articles/attachments`. The resizing was done with imagemagick:

`find docs/articles/attachments/ -maxdepth 2 -type f -exec mogrify -resize 300x200 {}\;`.

# Visualise data

## Datatable

```{r vis_data}
DT::datatable(fq_data)
DT::datatable(fq_data_tae)
```

## Map

Constructing label and popup requires knowledge of the dataset structure.

In absence of locally linked attachments (`get_attachment`), we reconstruct
the attachment URL, which (until ODK Central allows anonymous media retrieval) 
fails on the missing authentication.

```{r map_data}
leaflet::leaflet(width = 800, height = 600) %>%
  leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "Place names") %>%
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Aerial") %>%
  leaflet::clearBounds() %>% 
  leaflet::addAwesomeMarkers(
    data = fq_data,
    lng = ~...12, lat = ~...13,
    icon = leaflet::makeAwesomeIcon(text = "Q", markerColor = "red"),
    label = ~glue::glue('{area_name} {encounter_start_datetime}'),
    popup = ~glue::glue(
      "<h3>{area_name}</h3>",
      "Survey start {encounter_start_datetime}</br>",
      "Reporter {reporter}</br>",
      "Device {device_id}</br>",
      "<h5>Site</h5>",
      '<div><img src="{quadrat_photo}"',
      ' height="150px" alt="Quadrat photo"></img></div>',
      "<h5>Mudmap</h5>",
      '<div><img src="{mudmap_photo}',
      ' height="150px" alt="Mudmap"></img></div>',
      "<h5>Habitat</h5>",
      "Morphological type: {morphological_type}</br>",
      '<div><img src="{morphological_type_photo}"',
      'height="150px" alt="Morphological type"></img></div>',
      "Veg class: {vegclass_placeholder}</br>"
    ),
    clusterOptions = leaflet::markerClusterOptions()
  ) %>%
  leaflet::addLayersControl(
    baseGroups = c("Place names", "Aerial"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  )
```

```{r map_data_tae}
leaflet::leaflet(width = 800, height = 600) %>%
  leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "Place names") %>%
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Aerial") %>%
  leaflet::clearBounds() %>% 
  leaflet::addAwesomeMarkers(
    data = fq_data_tae,
    lng = ~...6, lat = ~...7,
    icon = leaflet::makeAwesomeIcon(text = "T", markerColor = "green"),
    label = ~glue::glue('{canonical_name} {encounter_start_datetime}'),
    popup = ~glue::glue(
      "<h3>{area_name}</h3>",
      "Survey start {encounter_start_datetime}</br>",
      "Reporter {reporter}</br>",
      "Device {device_id}</br>",
      "<h5>Taxon</h5>",
      '<div><img src="{photo_in_situ}"',
      ' height="150px" alt="Taxon in situ"></img></div>',
      "Specimen barcode: {voucher_specimen_barcode}</br>",
      "Life form: {life_form}</br>"
    ),
    clusterOptions = leaflet::markerClusterOptions()
  ) %>%
  leaflet::addLayersControl(
    baseGroups = c("Place names", "Aerial"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  )
```

