---
title: "Accessing the OData API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{odata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- media/1568786958640.jpg
- media/1568787004467.jpg
- media/1568787172983.jpg
- media/1568787220913.jpg
- media/1568787333689.jpg
- media/1568787380833.jpg
- media/1568787456664.jpg
- media/1568787504877.jpg
- media/1568794395624.jpg
- media/1568794560256.jpg
- media/1568795382633.jpg
- media/1568795579890.jpg
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(DT)
library(leaflet)
library(listviewer)
library(magrittr)
library(tibble)
library(tidyr)
library(lubridate)
library(knitr)
library(dplyr)
library(ggplot2)
library(ruODK)
```
<!-- TODO split out vignettes -->
<!-- * This vignette: the typical workflow -->
<!-- * New vignette: spatial is not special -->
<!-- * New vignette: DIY rectangling -->
<!-- * New vignette: data visualisation (refer to that from odata and resful api) -->
This vignette demonstrates `ruODK`'s workflow to extract data from ODK Central's
OData service endpoint, and to prepare the data and the media attachments for
further analysis and visualisation.

The demonstrated workflow is roughly equivalent to ODK Central's "Export all data",
which downloads all submissions and all repeating subgroups as CSV spreadsheets,
and all media attachments in a local subfolder "attachments".

An alternative pathway to getting data out of ODK Central is to use the REST API
as documented (with live examples in multiple programming languages) at the
[ODK Central API docs](https://odkcentral.docs.apiary.io/).

# Configure ruODK
The OData service URL is shown in the form's "Submissions" tab >
"Analyse via OData" on ODK Central. It contains base URL, project ID, and 
form ID and is used by `ruODK::ru_setup()`.

```{r ru_setup}
# ODK Central's OData URL contains base URL, project ID, and form ID
# ODK Central credentials can live in .Renviron, see vignette "setup"
ruODK::ru_setup(
  svc = paste0("https://sandbox.central.getodk.org/v1/projects/14/",
               "forms/build_Flora-Quadrat-0-4_1564384341.svc"),
  un = Sys.getenv("ODKC_TEST_UN"),
  pw = Sys.getenv("ODKC_TEST_PW"),
  tz = "Australia/Perth",
  verbose = TRUE
)

# File attachment download location
loc <- fs::path("media")

# Canned data
data("fq_svc")
data("fq_form_schema")
data("fq_meta")
data("fq_raw")
data("fq_raw_strata")
data("fq_raw_taxa")
data("fq_data")
data("fq_data_strata")
data("fq_data_taxa")
```

To extract data from the OData API endpoints, we have to:

* discover data endpoints from the OData service document,
* inspect the metadata schema to infer data types,
* download submissions from the data endpoints,
* download media attachments and adjust their file paths to the downloaded files.

## OData service document
Let's start off with the service document.

```{r load_odata_service, eval=F}
fq_svc <- ruODK::odata_service_get()
```

The same data is included as example data `fq_svc`.

```{r view_service}
fq_svc %>% knitr::kable(.)
```

`ruODK` provides the names and urls of the service endpoints as `tibble`.
We see the main data available under the url `Submissions`, and repeating 
form groups called `taxon_encounter` and `vegetation_stratum` in the ODK form under 
the url `Submissions.taxon_encounter` and `Submissions.vegetation_stratum`, 
respectively.

The main value we get out of the service document are these names of the 
form groups, which can differ between forms.

## OData metadata document
Let's inspect the form metadata to review our data schema.
While we can download the submission data without it, the metadata document 
contains information about field data types and attachment names.

```{r load_metadata, eval=FALSE}
fq_meta <- ruODK::odata_metadata_get()
```

```{r view_metadata, fig.width=7}
listviewer::jsonedit(fq_meta)
```

As an alternative to the OData metadata document, ODK Central also offers
form metadata as a much cleaner JSON document, which `ruODK` can read and parse
into a clean `tibble` of field type, name, and path. 

`ruODK` uses this introspection to parse submission data.

```{r form_schema, eval=FALSE}
fq_form_schema <- ruODK::form_schema()
```

```{r form_schema_show}
fq_form_schema %>% knitr::kable(.)
```

## OData submission data documents
Now let's download the form submissions and, separately, repeating form groups.
`ruODK::odata_submission_get()` defaults to download the submission data,
parse it into a tidy tibble, parses dates and datetimes, downloads
and links file attachments, and handles spatial datatypes.

This vignette is built with canned data, so the verbose messages are not shown.

With `wkt=TRUE`, we'll receive spatial types as Well Known Text, which `ruODK` parses
as plain text.
With `wkt=FALSE` (the default), we'll receive spatial types as GeoJSON, which
`ruODK` parses into a nested list.
`ruODK` retains the original spatial field, and annotates the data with 
extracted longitude, latitude, altitude, and (where given) accuracy.
These additional fields are prefixed with the original field name to prevent 
name collisions between possibly multiple location fields.

```{r load_odata, eval=FALSE}
fq_data <- ruODK::odata_submission_get(
  table = fq_svc$name[1], 
  local_dir = loc,
  wkt=TRUE)

fq_data_strata <- ruODK::odata_submission_get(
  table = fq_svc$name[2], 
  local_dir = loc
)

fq_data_taxa <- ruODK::odata_submission_get(
  table = fq_svc$name[3], 
  local_dir = loc,
  wkt=TRUE)
```

# Detour: Data rectangling
<!-- TODO: rewrite -->
<!-- * odata_submission_get(parse=F) -->
<!-- * odata_submission_rectangle(form_schema=fs) -->
<!-- * handle_ru_{datetimes, attachments, geopoints, geotraces, geoshapes} -->
The function `ruODK::odata_submission_get()` received the original XML response 
as a nested list of lists.
To analyse and visualise the data, this nested list of lists must be transformed 
into a rectangular shape.
The function `ruODK::odata_submission_rectangle()` is used internally to 
recursively un-nest list columns using `tidyr::unnest_wider()`.
Unnamed columns, notably the anonymous lat/lon/alt coordinates, are named 
automatically to become unique (a feature of `tidyr::unnest_*()`), and then 
sanitised using the helper `janitor::clean_names()`.

By default, form group names are used as prefix to the field names. 
This behaviour can be disabled by handing the argument `names_sep=NULL` to
`tidyr::unnest_wider()` through running 
`ruODK::odata_submission_get() %>% ruODK::odata_submission_rectangle(names_sep = NULL)`.

The vectorised function `ruODK::attachment_get()` is then used internally 
to download and link attachments like photos and other media to a local, 
relative path. This will take some time during the first run. 
Once the files exist locally, the download will be skipped.

When used through `ruODK::odata_submission_get()`, `ruODK` will introspect the
form schema to detect and then parse media attachment fields automatically.
When used manually, field names of media attachment fields can be (partially or 
fully) specified, see `??ruODK::attachment_get()`.

The date formats are parsed from ISO8601 timestamps into POSIXct objects with
`ruODK::handle_ru_datetimes()`. We use our local timezone (GMT+08) in this example.
`ruODK` introspects the form schema to detect and then parse date and datetime 
fields automatically.

The repeated subgroup `taxon_encounter` is left joined to the main submission data
to receive a (repeated) copy of the main submission data 
(such as location, time and habitat description). 
We will do the same to the other repeated subgroup `vegetation_stratum`.

For clarity, we enable verbose messages from `ruODK::odata_submission_get()` 
and preserve the message output in the code chunk options with `message=TRUE`.
In real-world use cases, messages can be disabled through the chunk option
`message=FALSE`.

We use a custom local path for attachments (`loc`). This results in a smaller
installed package size for `ruODK`, as it shares the attachment files with the
other vignettes. The default is a local folder `media`.

The raw and unparsed example data is provided as data objects 
`fq_raw` (main submissions of form Flora Quadrat 0.4), 
`fq_raw_taxa` (repeated group "Taxon Encounter" within a Flora Quadrat), and
`fq_raw_strata` (repeated group "Vegetation Stratum" within a Flora Quadrat).

The parsed versions are included as data objects `fq_data`, `fq_data_strata`, 
and `fq_data_taxa`. To enable users without ODK Central credentials to build
this vignette (e.g. on package installation with `build_vignettes=TRUE`),
we show the real functions (such as `ruODK::odata_submission_get()`), but do not
evaluate them. Instead, we use "canned data". The `ruODK` test suite ensures 
that canned data are equivalent to live data. 

The result of this code chunk should be exactly the same as the compact version
with `odata_submission_get(parse=TRUE)`.

```{r}
# Candidates for ruODK::handle_ru_datetimes()
fq_form_schema %>% 
  dplyr::filter(type %in% c("dateTime", "date")) %>% 
  knitr::kable(.)

# Candidates for ruODK::handle_ru_attachments()
fq_form_schema %>% 
  dplyr::filter(type == "binary") %>% 
  knitr::kable(.)

# Candidates for ruODK::handle_ru_geopoints()
fq_form_schema %>% 
  dplyr::filter(type == "geopoint") %>% 
  knitr::kable(.)
```


```{r, eval=FALSE}
# The raw submission data
fq_raw <- ruODK::odata_submission_get(table = fq_svc$name[1], parse = FALSE)
fq_strata <- ruODK::odata_submission_get(table = fq_svc$name[2], parse = FALSE)
fq_taxa <- ruODK::odata_submission_get(table = fq_svc$name[3], parse = FALSE)

# Parse main data
fq_data <- fq_raw %>%
  ruODK::odata_submission_rectangle() %>%
  ruODK::handle_ru_datetimes(fq_form_schema) %>%
  ruODK::handle_ru_geopoints(fq_form_schema) %>%
  ruODK::handle_ru_geotraces(fq_form_schema) %>%
  ruODK::handle_ru_geoshapes(fq_form_schema) %>% 
  ruODK::handle_ru_attachments(fq_form_schema, local_dir = t)

# Parse nested group "taxa"
fq_data_taxa <- fq_taxa %>% 
  ruODK::odata_submission_rectangle() %>%
  ruODK::handle_ru_datetimes(fq_form_schema) %>%
  ruODK::handle_ru_geopoints(fq_form_schema) %>%
  ruODK::handle_ru_geotraces(fq_form_schema) %>%
  ruODK::handle_ru_geoshapes(fq_form_schema) %>% 
  ruODK::handle_ru_attachments(fq_form_schema, local_dir = t) %>%
  dplyr::left_join(fq_data, by = c("submissions_id" = "id"))

# Parse nested group "strata"
fq_data_strata <- fq_strata %>% 
  ruODK::odata_submission_rectangle() %>%
  ruODK::handle_ru_datetimes(fq_form_schema) %>%
  ruODK::handle_ru_geopoints(fq_form_schema) %>%
  ruODK::handle_ru_geotraces(fq_form_schema) %>%
  ruODK::handle_ru_geoshapes(fq_form_schema) %>% 
  ruODK::handle_ru_attachments(fq_form_schema, local_dir = t) %>%
  dplyr::left_join(fq_data, by = c("submissions_id" = "id"))
```

Note: A manually resized version of the original photos in this example live in
the package source under [`articles/media`
](https://github.com/ropensci/ruODK/tree/gh-pages/articles/media). 
To minimise package size, they were resized with imagemagick:

```{sh, eval=FALSE}
find vignettes/media -type f -exec mogrify -resize 200x150 {} \\;
```

## DIY rectangling
For those wishing to go one step further, this section demonstrates the inner
workings of `ruODK`, the recursive use of `tidyr::unnest_wider()`.

The unnesting could also be done manually by building up a pipeline, which
stepwise unnests each list column.
This requires knowledge of the data structure, which can either be looked up
from the metadata, or by inspecting the raw data, `fq_raw`.

The following command has been built by stepwise adding `tidyr::unnest_wider()` 
expressions to the pipe until all list columns were eliminated.

The trailing `invisible()` allows us to toggle parts of the pipe by catching the
dangling `%>%`.

```{r rectangle_diy}
fq_data_diy <- tibble::tibble(value = fq_raw$value) %>%
  tidyr::unnest_wider(value) %>%
  # 1. find list columns:
  tidyr::unnest_wider(`__system`) %>%
  tidyr::unnest_wider(meta) %>%
  # add more lines here to unnest other form groups
  #
  # 2. rename column names
  dplyr::rename(
    uuid = `__id`
    # add more columns, e.g.
    # longitude=`...1`, latitude=`...2`, altitude=`...3`
  ) %>%
  # 3. handle media attachments
  # dplyr::mutate(photo_1 = attachment_get(data_url, uuid, photo_1)) %>%
  invisible()
```

# Visualise data
<!-- TODO: split out and show working examples for each vis (table, map, pivot, export) -->
<!-- for each data type (date, point, trace, shape) -->
This section provides some examples of standard data visualisations.

## Datatable
The package `DT` provides an interactive (and searchable) datatable.

```{r vis_data}
DT::datatable(fq_data)
DT::datatable(fq_data_taxa)
DT::datatable(fq_data_strata)
# DT::datatable(head(fq_data_diy))
```

## Map
The R package `leaflet` provides interactive maps.

Constructing label and popup requires knowledge of the dataset structure.

```{r map_data}
leaflet::leaflet(width = 800, height = 600) %>%
  leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "Place names") %>%
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Aerial") %>%
  leaflet::clearBounds() %>%
  leaflet::addAwesomeMarkers(
    data = fq_data,
    lng = ~location_corner1_longitude, 
    lat = ~location_corner1_latitude,
    icon = leaflet::makeAwesomeIcon(text = "Q", markerColor = "red"),
    label = ~ glue::glue("{location_area_name} {encounter_start_datetime}"),
    popup = ~ glue::glue(
      "<h3>{location_area_name}</h3>",
      "Survey start {encounter_start_datetime}</br>",
      "Reporter {reporter}</br>",
      "Device {device_id}</br>",
      "<h5>Site</h5>",
      '<div><img src="{location_quadrat_photo}"',
      ' height="150px" alt="Quadrat photo"></img></div>',
      "<h5>Mudmap</h5>",
      '<div><img src="{perimeter_mudmap_photo}',
      ' height="150px" alt="Mudmap"></img></div>',
      "<h5>Habitat</h5>",
      "Morphological type: {habitat_morphological_type}</br>",
      '<div><img src="{habitat_morphological_type_photo}"',
      'height="150px" alt="Morphological type"></img></div>'
    ),
    clusterOptions = leaflet::markerClusterOptions()
  ) %>%
  leaflet::addLayersControl(
    baseGroups = c("Place names", "Aerial"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  )
```

```{r map_data_tae}
leaflet::leaflet(width = 800, height = 600) %>%
  leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "Place names") %>%
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Aerial") %>%
  leaflet::clearBounds() %>%
  leaflet::addAwesomeMarkers(
    data = fq_data_taxa,
    lng = ~ location_corner1_longitude, 
    lat = ~ location_corner1_latitude,
    icon = leaflet::makeAwesomeIcon(text = "T", markerColor = "green"),
    label = ~ glue::glue("{field_name} {encounter_start_datetime}"),
    popup = ~ glue::glue(
      "<h3>{field_name}</h3>",
      "Survey start {encounter_start_datetime}</br>",
      "Reporter {reporter}</br>",
      "Device {device_id}</br>",
      "<h5>Taxon</h5>",
      '<div><img src="media/{photo_in_situ}"',
      ' height="150px" alt="Taxon in situ"></img></div>',
      "Specimen barcode: {voucher_specimen_barcode}</br>",
      "Life form: {life_form}</br>"
    ),
    clusterOptions = leaflet::markerClusterOptions()
  ) %>%
  leaflet::addLayersControl(
    baseGroups = c("Place names", "Aerial"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  )
```

## Summarising data
See Hadley Wickam's 
[R for Data Science](https://r4ds.had.co.nz/exploratory-data-analysis.html)
for more ideas on data exploration.

```{r eda, fig.height=5, fig.width=7}
# How many submissions per device?
fq_data %>% 
  dplyr::group_by(meta_instance_id) %>% 
  dplyr::tally() %>% 
  knitr::kable()

# How many species sightings per life form?
fq_data_taxa %>% 
  dplyr::group_by(life_form) %>% 
  dplyr::tally() %>% 
  knitr::kable()

# GGplot of a pivot table
fq_data_taxa  %>%
  dplyr::group_by(life_form) %>%
  dplyr::tally() %>%
  ggplot2::ggplot(ggplot2::aes(x = life_form, y = n)) +
  ggplot2::labs(
    title = "Title",
    subtitle = "Subtitle",
    x = "Life form",
    y = "Abundance"
  ) +
  ggplot2::geom_point() +
  ggplot2::theme_classic()

# GGplot with groups
fq_data_taxa %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x=encounter_start_datetime, 
      y=field_name, 
      colour=life_form,
      shape=meta_instance_id)) + 
  ggplot2::labs(
    title="Title", 
    subtitle="Subtitle", 
    x="Observation date", 
    y="Species", 
    colour="Life form",
    shape="Data collection device") +
  ggplot2::geom_point() + 
  ggplot2::theme_classic() + 
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Export
The rectangled data can now be exported. e.g. to CSV. Note that all list columns 
must be either unnested or dropped before exporting to CSV.

```{r export, eval=F}
fq_data %>% readr::write_csv("flora_quadrats.csv")
fq_data_tae %>% readr::write_csv("flora_quadrats_taxon_encounters.csv")
fq_data_veg %>% readr::write_csv("flora_quadrats_vegetation_strata.csv")
```

# ruReady to ODK?
In this vignette, we took a scenic tour through the general workflow of accessing
and wrangling ODK Central data using `ruODK`.

For your convenience, `ruODK` includes a template RMarkdown workbook with the
essential steps of the above workflow and colour-by-numbers instructions, which
can be used as a starting point for projects using data from ODK Central.

To create a new RMarkdown document from the `ruODK` template, 
run `rmarkdown::draft("test.Rmd", "odata", package="ruODK")`.

Users of RStudio can alternatively "Create a new RMarkdown document" 
"From template" and select `ruODK`'s template "ODK Central via OData".

Make sure to install a fresh version of `ruODK` to get the latest and greatest
template.
