---
title: "Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ruODK)
```

# Configure ruODK
Currently, ODK Central supports BasicAuth (username and password) and 
SessionAuth. `ruODK` supports only BasicAuth.

Each ruODK function accepts the server base URL, username, and password, with
defaults looking for R environment variables `ODKC_URL`, `ODKC_UN`, and `ODKC_PW`.

There are three ways of providing those settings to each function call.

## One-off: settings directly in function call

We can supply those credentials to each `ruODK` function call, but that's
cumbersome and not securely reproducible (the command would include the 
sensitive  credentials).

```{r, eval=FALSE}
ruODK::project_list(
  url="https://odkcentral.mydomain.com", 
  un="me@mail.com", 
  pw="......."
)
```


An example use case are the `ruODK` tests, which require settings different from
the settings used during development.

## Per session: environment variables

A better way is to set your ODK Central URL, username (email), and password as 
R environment variables.

`ruODK` functions default to use the R environment variables 
`ODKC_URL`, `ODKC_UN`, and `ODKC_PW` for authentication.
  
```{r, eval=F}
Sys.setenv(ODKC_URL="https://odkcentral.mydomain.com")
Sys.setenv(ODKC_UN="me@mail.com")
Sys.setenv(ODKC_PW=".......")
```

Now we can call `ruODK` functions without settings, and let `ruODK` fall back
to the environment variables:

```{r, eval=FALSE}
ruODK::project_list()
```


## Per machine: set and forget
Let's make those settings a permanent default.
Place them in your `~/.Renviron` to set them per machine,
or in a separate file outside of version control to set them per project.

Of course, there is a helper for editing `~/.Renviron` in `usethis`.

```{r, eval=FALSE}
usethis::edit_r_environ(scope = "user")
```

Note, the format for `~/.Renviron` is:
```{r, eval=FALSE}
ODKC_URL="https://odkcentral.dbca.wa.gov.au"
ODKC_UN="your@email.com"
ODKC_PW="..."
```


