---
title: "Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ruODK)
```

# Configure ruODK
`ruODK` functions work on a given ODK Central instance using a web user's 
credentials (username and password). Some functions also require project and
form IDs.

`ruODK`'s functions accept these parameters either as explicit keyword arguments,
or fall back to defaults.

`ruODK` suggests as best practice to set the defaults (or parts thereof) using 
`ruODK::ru_setup()` or through permanently setting environment variables.

## Best practice: `ru_setup`
`ruODK` provides helpers for settings, `ru_setup()` to set and `ru_settings()` 
to get settings.

While normal users of `ruODK` will only need a default pid, fid, url, username, 
and password, contributors to `ruODK` can include optional test server settings
which are required to run the test suite and build the vignettes.

`ruODK` infers the base URL, project and form ID from the OData Service URL
which is shown in ODK Central on the Form Submissions tab.

```{r ru_setup}
# ruODK user using OData service URL, username (an email), and password
ruODK::ru_setup(
  svc = "https://sandbox.central.opendatakit.org/v1/projects/14/forms/build_Flora-Quadrat-0-4_1564384341.svc", 
  un = "me@email.com", 
  pw = "..."
)

# ruODK contributor
ruODK::ru_setup(
  svc = "https://sandbox.central.opendatakit.org/v1/projects/14/forms/build_Flora-Quadrat-0-4_1564384341.svc", 
  un = "me@email.com", 
  pw = "my_default_password",
  test_svc = "https://sandbox.central.opendatakit.org/v1/projects/14/forms/build_Flora-Quadrat-0-4_1564384341.svc", 
  test_un = "me@email.com", 
  test_pw = "YYYY"
)

# ruODK contributor using base URL, project ID, and form ID
# Same effect as above, but more typing
ruODK::ru_setup(
  pid = 1,
  fid = "build_Flora-Quadrat-0-4_1564384341",
  url = "https://odkcentral.dbca.wa.gov.au", 
  un = "me@email.com", 
  pw = "my_default_password",
  test_url = "https://sandbox.central.opendatakit.org", 
  test_un = "me@email.com", 
  test_pw = "my_test_password",
  test_pid = 14,
  test_fid = "build_Flora-Quadrat-0-4_1564384341"
)

# Review settings
ruODK::ru_settings()
```

Now we can call `ruODK` functions without specifying `url`, `un`, and `pw`, 
and let `ruODK` fall back to the defaults:

```{r, eval=FALSE}
ruODK::project_list()
ruODK::submission_list()
```

## Permanent defaults: Environment variables
Read a great overview of R's startup process, and how environment variables
are sourced at the beginning of a new session at [whattheyforgot.org](https://whattheyforgot.org/r-startup.html).

`ruODK`'s functions default to the getters `ruODK::get_default_{pid,fid,url,un.pw}()`.
These getters in turn look up their values from environment variables.

A convenient way to have often used environment variables available is to add
them to `~/.Renviron` using `usethis::edit_r_environ(scope = "user")`.
This loads them freshly into a new session, eliminating the need to run 
`ru_setup()`. Note that the environment variables can be cleared or overwritten 
through calling `ru_setup()` or `Sys.setenv()` with respective arguments.

`ru_setup()` will not change any omitted arguments.

```{r open_renviron, eval=FALSE}
usethis::edit_r_environ(scope = "user")
```


```{r renviron, eval=FALSE}
ODKC_PID=1
ODKC_FID="build_Flora-Quadrat-0-2_1558575936"
ODKC_URL="https://odkcentral.dbca.wa.gov.au"
ODKC_UN="me@email.com"
ODKC_PW="..."
ODKC_TEST_URL="https://sandbox.central.opendatakit.org"
ODKC_TEST_UN="me@email.com"
ODKC_TEST_PW="..."
ODKC_TEST_PID=14
ODKC_TEST_FID="build_Flora-Quadrat-0-2_1558575936"
```


As an alternative to setting environment variables through `~/.Renviron`, 
you can set them through `Sys.setenv()`:

```{r setenv, eval=FALSE}
Sys.setenv(ODKC_URL="https://odkcentral.dbca.wa.gov.au")
Sys.setenv(ODKC_UN="me@mail.com")
Sys.setenv(ODKC_PW="...")
Sys.setenv(ODKC_TEST_URL="https://sandbox.central.opendatakit.org")
Sys.setenv(ODKC_TEST_UN="me@mail.com")
Sys.setenv(ODKC_TEST_PW="...")
Sys.setenv(ODKC_TEST_PID=14)
Sys.setenv(ODKC_TEST_FID="build_Flora-Quadrat-0-2_1558575936")
```



## The hard way: Per function call
We can supply those credentials to each `ruODK` function call.

```{r, eval=FALSE}
# Use ruODK without ru_setup
ruODK::project_list(
  url="https://my-odkc.com", 
  un="me@mail.com", 
  pw="......."
)

# Tests use default test settings explicitly
ruODK::project_list(
  url=ruODK::get_test_url(), 
  un=ruODK::get_test_un(), 
  pw=ruODK::get_test_pw()
)
```

An example use case are the `ruODK` tests, which explicitly set `url`, `un`, 
`pw`, `pid` and `fid` from the test variables `ruODK::get_test_{url, un, pw, pid, fid}()`.


# Moving across forms
`ruODK`'s functions default to the default values for project ID (`pid`),
form ID (`fid`), base URL (`url`), username (`un`), and password (`pw`).

A typical workflow is to run several functions of `ruODK` against one form (and
the overarching project). By running `ru_setup()` with the form's OData service
URL and a web user's username and password, subsequent functions can omit
`pid`, `fid`, `url`, `un`, and `pw`.

```{r, eval=FALSE}
# Server 1, Project 1, Form 1
ruODK::ru_setup(
  svc = "https://central1.org/v1/projects/1/forms/form1.svc",
  un = "me@email.com",
  pw = "..."
)

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 1, Project 1, Form 3
ruODK::ru_setup(svc = "https://central1.org/v1/projects/1/forms/form3.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 1, Project 5, Form 4
ruODK::ru_setup(svc = "https://central1.org/v1/projects/5/forms/form4.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 2, Project 11, Form 12
ruODK::ru_setup(
  svc = "https://central2.org/v1/projects/11/forms/form12.svc",
  un = "me@email.com",
  pw = "..."
)

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 2, Project 11, Form 13
ruODK::ru_setup(svc = "https://central2.org/v1/projects/11/forms/form13.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()
```
