---
title: "Setup"
output: 
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ruODK)
```

# Configure ruODK

`ruODK` functions work on a given ODK Central instance using a web user's
credentials (username and password). Some functions also require project and
form IDs.

`ruODK`'s functions accept these parameters either as explicit keyword
arguments, or fall back to defaults.

Note: Always consider code to be public. Never use plain text credentials in
code.

`ruODK` suggests as best practice to set the defaults (or parts thereof) using
`ruODK::ru_setup()` or through permanently setting environment variables.

## Best practice: `ru_setup`

`ruODK` provides helpers for settings, `ru_setup()` to set and `ru_settings()`
to get settings.

While normal users of `ruODK` will only need a default pid, fid, url, username,
and password, contributors to `ruODK` can include optional test server settings
which are required to run the test suite and build the vignettes.

`ruODK` infers the base URL, project and form ID from the OData Service URL
which is shown in ODK Central on the Form Submissions tab.

Unless specified as function parameter, `ruODK` converts dates and times to the
default timezone, in this example set to "Australia/Perth".

Furthermore, some functions offer verbose messages, which can assist to debug
unexpected behaviour. Unless specified in the settings, or in the respective
function calls, `ruODK` defaults to hide verbose messages.

```{r ru_setup, eval=FALSE}
# ruODK user using OData service URL, username (an email), and password
# Never use plaintext username and password, use Sys.getenv() instead
ruODK::ru_setup(
  svc = Sys.getenv("ODKC_SVC"), 
  un = Sys.getenv("ODKC_UN"), 
  pw = Sys.getenv("ODKC_PW"),
  tz = "Australia/Perth",
  verbose = TRUE
)

# ruODK contributors: see contributing guidelines for .Renviron variables

# Review settings
ruODK::ru_settings()
```

Now we can call `ruODK` functions without specifying `url`, `un`, and `pw`, and
let `ruODK` fall back to the defaults:

```{r, eval=FALSE}
ruODK::project_list()
ruODK::submission_list()
```

## Permanent defaults: Environment variables

Read a great overview of R's startup process, and how environment variables are
sourced at the beginning of a new session at
[whattheyforgot.org](https://whattheyforgot.org/r-startup.html).

`ruODK`'s functions default to the getters
`ruODK::get_default_{pid,fid,url,un.pw}()`. These getters in turn look up their
values from environment variables.

A convenient way to have often used environment variables available is to add
them to `~/.Renviron` using `usethis::edit_r_environ(scope = "user")`. This
loads them freshly into a new session, eliminating the need to run `ru_setup()`.
Note that the environment variables can be cleared or overwritten through
calling `ru_setup()` or `Sys.setenv()` with respective arguments.

`ru_setup()` will not change any omitted arguments.

```{r open_renviron, eval=FALSE}
usethis::edit_r_environ(scope = "user")
```

```{r renviron, eval=FALSE}
ODKC_PID=1
ODKC_FID="build_Flora-Quadrat-0-2_1558575936"
ODKC_URL="https://odkcentral.dbca.wa.gov.au"
ODKC_UN="me@email.com"
ODKC_PW="..."
ODKC_VERSION=0.8

# Test settings
ODKC_TEST_SVC="https://odkc.dbca.wa.gov.au/v1/projects/2/forms/Flora-Quadrat-04.svc"
ODKC_TEST_URL="https://odkc.dbca.wa.gov.au"
ODKC_TEST_PID=2
ODKC_TEST_PID_ENC=3
ODKC_TEST_PP="ThePassphrase"
ODKC_TEST_FID="Flora-Quadrat-04"
ODKC_TEST_FID_ZIP="Spotlighting-06"
ODKC_TEST_FID_ATT="Flora-Quadrat-04-att"
ODKC_TEST_FID_GAP="Flora-Quadrat-04-gap"
ODKC_TEST_FID_WKT="Locations"
ODKC_TEST_FID_I8N0="I8n_no_lang"
ODKC_TEST_FID_I8N1="I8n_label_lng"
ODKC_TEST_FID_I8N2="I8n_label_choices"
ODKC_TEST_FID_I8N3="I8n_no_lang_choicefilter"
ODKC_TEST_FID_I8N4="I8n_lang_choicefilter"
ODKC_TEST_FID_ENC="Locations"
ODKC_TEST_VERSION=1.0
RU_VERBOSE=TRUE
RU_TIMEZONE="Australia/Perth"
RU_RETRIES=3
ODKC_TEST_UN="..."
ODKC_TEST_PW="..."
```

As an alternative to setting environment variables through `~/.Renviron`, you
can set them through `Sys.setenv()`:

```{r setenv, eval=FALSE}
Sys.setenv(ODKC_URL="https://odkc.dbca.wa.gov.au")
Sys.setenv(ODKC_UN="me@mail.com")
Sys.setenv(ODKC_PW="...")
Sys.setenv(ODKC_TEST_URL="...")
Sys.setenv(ODKC_TEST_UN="...")
Sys.setenv(ODKC_TEST_PW="...")
Sys.setenv(ODKC_TEST_PID=2)
# continue for the remaining test variables shown above
```

## The hard way: Per function call

We can supply those credentials to each `ruODK` function call.

Note, including sensitive credentials in plain text into code is bad practice.
This option is shown only for completeness.

```{r, eval=FALSE}
# Use ruODK without ru_setup
ruODK::project_list(
  url="https://my-odkc.com", 
  un = Sys.getenv("ODKC_UN"), 
  pw = Sys.getenv("ODKC_PW")
)

# Tests use default test settings explicitly
ruODK::project_list(
  url=ruODK::get_test_url(),
  un=ruODK::get_test_un(), 
  pw=ruODK::get_test_pw()
)
```

An example use case are the `ruODK` tests, which explicitly set `url`, `un`,
`pw`, `pid` and `fid` from the test variables
`ruODK::get_test_{url, un, pw, pid, fid}()`. Note that this uses functions
instead of plain text versions of sensitive credentials. Alternatively,
variables could also be used to set credentials per function call.

## Moving across forms

`ruODK`'s functions default to the default values for project ID (`pid`), form
ID (`fid`), base URL (`url`), username (`un`), and password (`pw`).

A typical workflow is to run several functions of `ruODK` against one form (and
the overarching project). By running `ru_setup()` with the form's OData service
URL and a web user's username and password, subsequent functions can omit `pid`,
`fid`, `url`, `un`, and `pw`.

```{r, eval=FALSE}
# Server 1, Project 1, Form 1
ruODK::ru_setup(
  svc = "https://central1.org/v1/projects/1/forms/form1.svc",
  un = Sys.getenv("ODKC_UN"), 
  pw = Sys.getenv("ODKC_PW")
)

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 1, Project 1, Form 3
ruODK::ru_setup(svc = "https://central1.org/v1/projects/1/forms/form3.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 1, Project 5, Form 4
ruODK::ru_setup(svc = "https://central1.org/v1/projects/5/forms/form4.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 2, Project 11, Form 12
ruODK::ru_setup(
  svc = "https://central2.org/v1/projects/11/forms/form12.svc",
  un = Sys.getenv("ODKC_UN2"), 
  pw = Sys.getenv("ODKC_PW2")
)

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()

# Server 2, Project 11, Form 13
ruODK::ru_setup(svc = "https://central2.org/v1/projects/11/forms/form13.svc")

ruODK::project_detail()
ruODK::form_detail()
ruODK::submission_list()
```

## Legacy support for deprecated ODK Central features

Users of the latest ODK Central version can safely ignore this section. Users of
older ODK Central versions will be exposed to breaking changes coming from
upstream ODK Central.

Very occasionally, ODK Central will deprecate API endpoints. `ruODK` aims to
keep up with the latest ODK Central API, and will adapt its behaviour to support
both new and old, deprecated behaviour.

As there is no corresponding API endpoint to determine the ODK Central version,
`ruODK` has introduced the environment variables `ODKC_VERSION` and
`ODKC_TEST_VERSION`. If unset, `ruODK` defaults to the latest ODK Central
version, which is the version deployed to the ODK Central sandbox (on which
`ruODK`'s package tests run).

The numeric format of `major.minor` (e.g. `0.7` or `0.8`) future-proofs `ruODK`
to be able to handle any other future breaking changes gracefully through the
same variable.

A recent example is the change of the API endpoint for `form_schema`, which
replaced the `.schema.json` API in favour of the `/fields` API. This required
`ruODK` to toggle its behaviour between the relatively involved un-nesting of
the JSON schema (versions before 0.8) with parsing a clean, flat list of field
names, types, and XForms paths (version 0.8 onward).
