---
title: "Spatial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

For forms with spatial types, such as geopoint, geotrace, or geoshape, 
`ruODK` gives two options to access the captured spatial data.

Firstly, to make spatial data as simple and accessible as possible, 
`ruODK` extracts the lat/lon/alt/acc from geopoints, as well as from 
the first coordinate of geotraces and geoshapes into separate columns.
This works for both GeoJSON and WKT.
The extracted columns are named as the original geofield appended with 
`_latitude`, `_longitude`, `_altitude`, and `_accuracy`, respectively.

Secondly, this vignette demonstrates how to turn the spatial data types returned 
from ODK Central into spatially enabled objects. 
To do so, we have to address two challenges.

The first challenge is to select which of the potentially many spatial fields
which an ODK form can capture shall be used as the primary geometry of a native
spatial object, such as an `sf` SimpleFeature class.
If several spatial fields are captured, it is up to the user to choose which 
field to use as primary geometry.

The second difficulty is that the parsed data from ODK Central is a plain table 
(`tbl_df`) in which some columns contain spatial data. WKT is stored in text 
columns, GeoJSON is stored in list columns. This format is not recognised as a
standard spatial input format.

```{r, message=FALSE}
library(ruODK)

# Visualisation
library(leaflet)
library(ggplot2)
library(mapview)

# Spatial
library(sf)
```

## Data

The data shown in this vignette is hosted on the ODK Central Sandbox as one of 
the forms used in `ruODK` package tests. The form contains every spatial widget
supported by ODK Build for every supported spatial field type.

With working credentials to the ODK Sandbox, we can download the data directly.

```{r dl_submissions, eval=FALSE}
# Set ruODK defaults to an ODK Central form, choose tz and verbosity
ruODK::ru_setup(
  svc=paste0("https://sandbox.central.getodk.org/v1/projects/14/",
             "forms/build_Locations_1589344221.svc"),
  un = get_test_un(), 
  pw = get_test_pw(),
  odkc_version = 0.8,
  tz = "Australia/Perth",
  verbose = TRUE)
data_wkt <- ruODK::odata_submission_get(wkt = TRUE)
data_gj <- ruODK::odata_submission_get(wkt = FALSE)
```

To allow users to build this vignette without credentials to the ODK Central 
Sandbox, `ruODK` provides above form data also as package data. 

```{r}
data("geo_wkt", package = "ruODK")
data("geo_gj", package = "ruODK")
```


## Map geopoints
We can turn data with a text column containing WKT into a SimpleFeatures object.

In addition, we can leave the `tbl_df` as non-spatial object, and instead 
use the separately extracted latitude, 
longitude, altitude, and accuracy individually e.g. to plot a Leaflet map.

<!--  
mapview's popup is temporarily broken:
  https://github.com/r-spatial/mapview/issues/312

We disable the popup with `popup = NULL` until the mapview fix hits CRAN.
-->

```{r, fig.width=9, fig.height=7}
geo_sf_point <- geo_wkt %>% sf::st_as_sf(wkt="point_location_point_gps")
# Mapview using sf
# Remove the popup=NULL to get full attribute table as popup
# See also mention under heading "Outlook"
mapview::mapview(geo_sf_point, col.regions = sf::sf.colors(10))
# GGplot using sf
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = geo_sf_point, ggplot2::aes(fill = device_id))
# Leaflet using sf
leaflet::leaflet(data = geo_sf_point) %>% 
  leaflet::addTiles() %>% 
  leaflet::addMarkers(label = ~ device_id, popup = ~ device_id)
# Leflet using extracted coordinate components in tbl_df
leaflet::leaflet(data = geo_wkt) %>% 
  leaflet::addTiles() %>% 
  leaflet::addMarkers(
    lng = ~ point_location_point_gps_longitude,
    lat = ~ point_location_point_gps_latitude,
    label = ~ device_id, 
    popup = ~ device_id)
```


## Map geotraces (lines)
We use `sf::st_as_sf` on a text column containing a WKT geotrace.

```{r, fig.width=9, fig.height=7}
geo_sf_line <- geo_wkt %>% sf::st_as_sf(wkt="path_location_path_gps")
# Mapview using sf
mapview::mapview(geo_sf_line, col.regions = sf::sf.colors(10), popup = NULL)
# GGplot using sf
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = geo_sf_line, ggplot2::aes(fill = device_id))
# Leaflet using first extracted coordinate compoments from plain tbl_df
# Note this represents the geotrace (line) with a simple point
leaflet::leaflet(data = geo_wkt) %>% 
  leaflet::addTiles() %>% 
  leaflet::addMarkers(
    lng = ~ path_location_path_gps_longitude,
    lat = ~ path_location_path_gps_latitude,
    label = ~ device_id, 
    popup = ~ device_id)
```

## Map geoshapes (polygons)
Again, we'll use `sf::st_as_sf` but select a WKT geoshape column.

```{r, fig.width=9, fig.height=7}
geo_sf_poly <- geo_wkt %>% sf::st_as_sf(wkt="shape_location_shape_gps")
# Mapview using sf
mapview::mapview(geo_sf_poly, col.regions = sf::sf.colors(10), popup = NULL)
# GGplot using sf
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = geo_sf_poly, ggplot2::aes(fill = device_id))
# Leaflet using first extracted coordinate compoments from plain tbl_df
# Note this represents the geoshape (polygon) with a simple point
leaflet::leaflet(data = geo_wkt) %>% 
  leaflet::addTiles() %>% 
  leaflet::addMarkers(
    lng = ~ shape_location_shape_gps_longitude,
    lat = ~ shape_location_shape_gps_latitude,
    label = ~ device_id, 
    popup = ~ device_id)
```

# Outlook
The above examples show how to turn spatial data into an `sf` object, and give
very rudimentary visualisation examples.

See the [sf homepage](https://r-spatial.github.io/sf/) for more
context and examples. 
The [sf cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/sf.pdf) 
deserves a spatial mention.
Review the options for 
[mapview popups](https://r-spatial.github.io/mapview/articles/articles/mapview_04-popups.html) 
which are turned off in this quick demo.

There are several other good entry points for all things R and spatial, 
including but not limited to:

* The [R Spatial CRAN Task View](https://cran.r-project.org/web/views/Spatial.html)
* The [RSpatial](https://rspatial.org/) website
* [Geospatial data in R and beyond](https://www.maths.lancs.ac.uk/~rowlings/Teaching/UseR2012/) 
  by [Barry Rowlingson](http://barry.rowlingson.com/)
* [GIS with R](https://www.jessesadler.com/post/gis-with-r-intro/) 
  by [Jesse Sadler](https://www.jessesadler.com/page/cv/)
  
The above list of examples and resources is far from comprehensive.
Feel free to [contribute or suggest](https://github.com/ropensci/ruODK/issues) 
other working examples for turning data from `ruODK` into spatial formats.
